Paleoenvironmental models for Australia and the impact of aridification on blindsnake diversification

### 2023-05-16 manuscript.Rmd
update to match word manuscript after revision

**Running title** Decrease in blindsnake speciation correlate with aridification

**Authors**: Sarin Tiatragul^1^; Alexander Skeels^2,3^; J. Scott Keogh^1^

@@

# Acknowledgements
ST would like to thank J Rolland and F. Condamine for helpful advice about fitting diversification models with RPANDA. Thanks to NJ Matzke and O Jiménez-Robles for help with BioGeoBEARS. Special thanks to IG Brennan, D Esquerré, M Pepper, CJ Pavón-Vázquez, and O Hagen for helpful advice on the analysis and discussion. JSK thanks the Australian Research Council for ongoing support.

@@

Today, the arid zone covers roughly 70% of the Australian landscape and provides habitat to 43% of all of Australia’s terrestrial reptiles [@byrneBirth2008].Of these, roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus *Anilios*, which currently comprises 47 described species  **47** described species [@uetzQuarter2021] (Tiatragul et al., submitted MS). All but three species of currenty described _Anilios_ are endemic to Australia, _A. torresianus_ and _A. leucoproctus_ also occur in southern parts of New Guinea, while _A. eryncinus_ has only been recorded in New Guinea and the Maluku Islands. 
Blindsnakes are morphologically conserved and previous phylogenetic and morphological studies have revealed that the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019]. The diversification of _Anilios_ in Australia is reported to have originated from a single dispersal event by a mesic-adapted ancestor from Southeast Asia or Indonesia between 39 – 19 Ma [@vidalBlindsnake2010;@marinTracing2013]. 
However, phylogenetic analyses from Tiatragul et al. (2023) indicated that the closest relative to all *Anilios* is *Ramphotyphlops multilineatus*, a species that occur in Western New Guinea, Kai Islands, and Salawati, suggesting the Australian radiation may have a Melanesian origin
Despite representing a substantial portion of snake diversity, we know little about the macroevolutionary dynamics of blindsnakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic environmental changes during the Cenozoic because they are geographically restricted and can be found in every biome on mainland Australia.

@@

We used the most updated phylogeny for *Anilios* and closely related outgroups from Tiatragul et al., submitted MS. The species tree was built with the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] using maximum likelihood gene trees inferred in IQTREE from 4930 loci (322 Anchored Hybrid Enrichment, 4574 ultraconserved elements, and 34 legacy genes) for 35/47 recognised *Anilios* species. Tiatragul et al. (submitted) time-calibrated the phylogeny using 27 outgroup squamate fossils as primary calibrations using MCMCTree from the PAML package [@yangPAML2007] (see Tiatragul et al., submitted MS for details). Phylogenetic data in Tiatragul et al. (submitted) suggest the existence of at least three genetically distinct undescribed species, so we followed their designation and adjusted our taxon sampling fraction to 38/50 species. 
The genetically distinct OTUs that are not fully described species include three '_Anilios grypus_' (two in Western Australia and one in Queensland) and two '_Anilios ligatus_' (one from the eastern part of the range and one from the northern part) (Tiatragul et al., submitted MS). 
For ancestral biome estimation analyses, we included three outgroup blindsnakes species from Southeast Asia, Indonesia, and Melanesia, where the ancestor of *Anilios* has been hypothesised to have originated [@vidalBlindsnake2010]. The three species include *Ramphotyphlops multilineatus* (Western New Guinea, Kai Island, and Salawati), *Sundatyphlops polygrammicus* (Moyo Island, Indonesia), and *Acutotyphlops subocularis* (Bismarck Archipelago). We were not able to include a *Ramphotyphlops* from continental Southeast Asia in the time-calibrated phylogeny because insufficient data was available for divergence time estimation.

@@

To investigate the biogeographic history of blindsnakes, we used the R package 'BioGeoBEARS' v.1.1.2 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible, likelihood-based framework that models how branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events, including processes such as dispersal and extinction) or at the time of branching (cladogenetic events involving sympatric, vicariant, and founder-event speciation). 
We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records from the Atlas of Living Australia database. Biome classifications were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World data set [@olsonTerrestrial2001] and followed the nomenclature of published Australia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021].

@@

To test whether diversification rates differ between arid-adapted and mesic-adapted lineages, we used two approaches. First, we estimated speciation rates at the origin of each branch (henceforth branch-specific rates) using ClaDS function in the Julia package 'PANDA' v.1.6.1 [@malietFast2022]. ClaDS is designed to capture frequent shifts in speciation rates with small effects. At each speciation event, daughter lineages are assigned its own speciation rate sampled from a distribution parameterised on the parental rate [@malietFast2022].

@@

Based on current distributions from ALA, most Anilios species are found entirely or partially in the tropical grassland biome (n = 20) followed by the arid biome (n = 13) and Mediterranean biome (n = 10). Apart from A. silvia, A. leucoproctus, and A. proximus that appeared to form independent divergent lineages (Fig. \@ref(fig:fig-biogeobears)a), we identified divergent biogeographic transitions between two major groups of species, which we refer to as Group 1 (n = 21 lineages) and Group 2 (n = 14 lineages). The common ancestor of Group 1 remained mostly in the tropical grassland biome, then 12/21 lineages expanded or transitioned into other biomes. The common ancestor of Group 2 first transitioned into the arid biome, four lineages remained in the arid biome while ten lineages transitioned or expanded into other biomes.

The tropical grassland biome is the largest source of outward dispersal (16.2 ± 4.9 events; 46.6%) followed by the arid biome (9.7 ± 3.2 events; 26%)(Fig. 1b). On the other hand, the largest sinks for all inward dispersals are the tropical forest biome (7.9 ± 4.2 events; 21.2%) and the Mediterranean biome (7.0 ± 2.9 events; 18.8%). The greatest directional transition is from the arid biome to the Mediterranean biome (5.3 ± 1.5 events) and from the tropical grassland biome into the temperate forest biome (5.1 ± 1.3 events) (Fig. \@ref(fig:fig-biogeobears)c).

```{r fig-biogeobears, fig.cap= 'Ancestral biome estimation and transitions of Australo-Papuan blindsnakes in BioGeoBEARS, using the DEC+j+x model. (a) Pie charts based on 50 biogeographic stochastic map (BSM) representing probabilities of each biome at speciation event. Inset map (left) shows the current distribution of biomes. Inset maps to the right of the phylogeny shows distribution of species in Group 1 and Group 2. (b) Bar plot showing frequency and composition of outward transitions and (c) inward transitions from Australian biomes and combined non-Australian islands. Abbreviations: TrG = Tropical Grassland, TmF = Temperate Forest, AR = Arid, MD = Mediterranean, TrF = Tropical Forest, and TmG = Temperate Grassland.'}

@@

```{r fig-paleoclimate-compare, fig.cap= 'Paleoclimate estimation based on three different methods and RPANDA. (a) Estimated mean temperature over time in Australia compared to RPANDA global paleotemperature data. (b) Estimated mean aridity index over time based on precipitation threshold. Abbreviations Pli = Pliocene and Ple = Pleistocene.'}

@@

```{r fig-RPANDA, fig.cap= 'Effect of aridity change on the diversification of Anilios estimated from the phylogeny. (a) The aridity index through time calculated from reconstructed paleoclimate (dotted line), and lineage through time estimated from the phylogeny (solid line). (b) Fitted diversification rates through time reconstructed from parameter estimates from the best diversification model (speciation rates vary exponentially with paleoaridity and constant extinction rate; “BEnvVar + DCST (EXPO)”). (c) Fitted diversification rates at different aridity index values calculated from parameter estimates of the previous model. Increased aridity is associated with decrease net diversification rates in this model..'}

@@

Tiatragul et al., (submitted) proposed that the earliest divergence of *Anilios* occurred around 25 Ma (28 – 22 Ma; 95% HPD), which is slightly older than 21.9 Ma (30.2 – 15.1 Ma; 95% Bayesian credibility interval) proposed by previous studies [@marinTracing2013;@vidalBlindsnake2010]. Our BioGeoBEARS analyses suggest that _Anilios_ gradually diversified in the tropical grassland (Group 1) and arid biome (Group 2) during a time when Australia was relatively wet and humid [@byrneBirth2008]. This indicates that many species found in arid biomes, particularly in Group 2, diversified in the arid biome prior to the onset of widespread aridification in the late Miocene [@byrneBirth2008].

@@

## Limitations
Although we did not detect conflicting signal between our analyses, some of our results, particularly those that estimate diversification rates (GeoHiSSE, ClaDS, and RPANDA), may be biased due to the small sample size of our _Anilios_ phylogeny [@caetanoHidden2018;@malietModel2019;@sunEstimating2020]. However, this near-complete phylogeny of _Anilios_ is the most comprehensive data set available at this time. 
We suspect there remains several undescribed species of blindsnakes in Australia and New Guinea that may belong in the genus _Anilios_. Getting genetic data from _A. erycinus_ will allow us to place the species among other _Anilios_ and provide a more comprehensive picture of the evolution of this genus.

>>>>

**Running title** Blindsnake speciation slows down with aridification

**Authors**: Sarin Tiatragul^1^; Alexander Skeels1^,^2,3^; J. Scott Keogh^1^

@@

# Acknowledgements
ST would like to thank J Rolland and F. Condamine for their helpful advice about fitting diversification models with RPANDA. Thanks to NJ Matzke and O Jiménez-Robles for help with BioGeoBEARS. Special thanks to IG Brennan, D Esquerré, M Pepper, CJ Pavón-Vázquez, and O Hagen for helpful advice on the analyses and discussion. JSK thanks the Australian Research Council for ongoing support.

@@

Today, the arid zone covers roughly 70% of the Australian landscape and provides habitat to 43% of all of Australia’s terrestrial reptiles [@byrneBirth2008].Of these, roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus *Anilios*, which currently comprises 47 described species **47** described species [@uetzQuarter2021] (Tiatragul et al., accepted). All but three species of currenty described _Anilios_ are endemic to Australia, _A. torresianus_ and _A. leucoproctus_ also occur in southern parts of New Guinea, while _A. eryncinus_ has only been recorded in New Guinea and the Maluku Islands. 
Blindsnakes are morphologically conserved and previous phylogenetic and morphological studies have revealed that the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019]. The diversification of _Anilios_ in Australia is reported to have originated from a single dispersal event by a mesic-adapted ancestor from Southeast Asia or Indonesia between 39 – 19 Ma [@vidalBlindsnake2010;@marinTracing2013]. 
However, phylogenetic analyses from Tiatragul et al. (accepted) indicated that the closest relative to all *Anilios* is *Ramphotyphlops multilineatus*, a species that occur in Western New Guinea, Kai Islands, and Salawati, suggesting the Australian radiation may have a Melanesian origin.
Despite representing a substantial portion of snake diversity, we know little about the macroevolutionary dynamics of blindsnakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic environmental changes during the Cenozoic because they are geographically restricted and can be found in every biome on mainland Australia.

@@

We used the most updated phylogeny for *Anilios* and closely related outgroups from Tiatragul et al., submitted MS. The species tree was built with the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] using maximum likelihood gene trees inferred in IQTREE from 4930 loci (322 Anchored Hybrid Enrichment, 4574 ultraconserved elements, and 34 legacy genes) for 35/47 recognised *Anilios* species. Tiatragul et al. (accepted) time-calibrated the phylogeny using 27 outgroup squamate fossils as primary calibrations using MCMCTree from the PAML package [@yangPAML2007]. Phylogenetic data in Tiatragul et al. (submitted) suggest the existence of at least three genetically distinct undescribed species, so we followed their designation and adjusted our taxon sampling fraction to 38/50 species. 
The genetically distinct OTUs that are not fully described species include three '_Anilios grypus_' (two in Western Australia and one in Queensland) and two '_Anilios ligatus_' (one from the eastern part of the range and one from the northern part) (Tiatragul et al., accepted). 
For ancestral biome estimation analyses, we included three outgroup blindsnakes species from Southeast Asia, Indonesia, and Melanesia, where the ancestor of *Anilios* has been hypothesised to have originated [@vidalBlindsnake2010; Tiatragul et al., accepted]. The three species include *Ramphotyphlops multilineatus* (Western New Guinea, Kai Island, and Salawati), *Sundatyphlops polygrammicus* (Moyo Island, Indonesia), and *Acutotyphlops subocularis* (Bismarck Archipelago). We were not able to include a *Ramphotyphlops* from continental Southeast Asia in the time-calibrated phylogeny because insufficient data was available for divergence time estimation.

@@

To investigate the biogeographic history of blindsnakes, we used the R package 'BioGeoBEARS' v.1.1.2 [@matzkeBioGeoBEARS2013]. BioGeoBEARS provides a flexible, likelihood-based framework that models how branches in the phylogeny (representing extant taxa and their ancestors) have shifted among a set of discrete biomes over time. These shifts in biomes can occur within a single branch (anagenetic events, including processes such as dispersal and extinction) or at the time of branching (cladogenetic events involving sympatric, vicariant, and founder-event speciation). 
We divided Australo-Papua into nine discrete biomes and assigned each Australian species to one or more biomes based on their occurrence records from the Atlas of Living Australia database (ALA). Biome classifications were based on the World Wildlife Fund's (WWF) Terrestrial Ecoregions of the World data set [@olsonTerrestrial2001] and followed the nomenclature of published Australia-specific studies [@cardilloPhylogeny2017;@pavon-vazquezComprehensive2021].

@@

To test whether diversification rates differ between arid-adapted and mesic-adapted lineages, we used two approaches. First, we estimated speciation rates at the origin of each branch (henceforth branch-specific rates) using ClaDS function in the Julia package PANDA v.1.6.1 [@malietFast2022]. ClaDS is designed to capture frequent shifts in speciation rates with small effects. At each speciation event, daughter lineages are assigned its own speciation rate sampled from a distribution parameterised on the parental rate [@malietFast2022].

@@

Based on current distributions from ALA, most *Anilios* species are found entirely or partially in the tropical grassland biome (n = 20) followed by the arid biome (n = 13) and Mediterranean biome (n = 10). Apart from *A. silvia*, *A. leucoproctus*, and *A. proximus* that appeared to form independent divergent lineages (Fig. \@ref(fig:fig-biogeobears)a), we identified divergent biogeographic transitions between two major groups of species, which we refer to as Group 1 (n = 21 lineages) and Group 2 (n = 14 lineages). The common ancestor of Group 1 remained mostly in the tropical grassland biome, then 12/21 lineages expanded or transitioned into other biomes. The common ancestor of Group 2 first transitioned into the arid biome, four lineages remained in the arid biome while ten lineages transitioned or expanded into other biomes.

The tropical grassland biome is the largest source of outward dispersal (16.2 ± 4.9 events; 46.6%) followed by the arid biome (9.7 ± 3.2 events; 26%)(Fig. \@ref(fig:fig-biogeobears)b). On the other hand, the largest sinks for all inward dispersals are the tropical forest biome (7.9 ± 4.2 events; 21.2%) and the Mediterranean biome (7.0 ± 2.9 events; 18.8%). The greatest directional transition is from the arid biome to the Mediterranean biome (5.3 ± 1.5 events) and from the tropical grassland biome into the temperate forest biome (5.1 ± 1.3 events) (Fig. \@ref(fig:fig-biogeobears)c).

```{r fig-biogeobears, fig.cap= 'Ancestral biome estimation and transitions of Australo-Papuan blindsnakes in BioGeoBEARS, using the DEC+j+x model. (a) Inset map (left) shows biome classifications of the regions considered in this study. Coloured boxes at the tips correspond to biomes on the map. Pie charts represent ancestral biome reconstruction based on 50 biogeographic stochastic maps (BSM) for each node. Colours present on the pie charts but not in the biome map indicate mixture of more than one biome. Coloured boxes below pie charts indicate the most probable biome(s) at visually ambiguous nodes. Inset maps (right of the phylogeny) shows distribution of species in Group 1 and Group 2. (b) Bar plot showing frequency and composition of outward transitions and (c) inward transitions from Australian biomes and combined non-Australian islands. Abbreviations: TrG = Tropical Grassland, TmF = Temperate Forest, AR = Arid, MD = Mediterranean, TrF = Tropical Forest, and TmG = Temperate Grassland.'}

@@

```{r fig-paleoclimate-compare, fig.cap= 'Paleoclimate estimation derived from RPANDA (pink),  a global circulation model (GCM) approach (black = Valdes et al., 2021), and two paleo-Köppen climate bands approach (orange = Straume et al., 2020; blue = Scotese & Wright, 2018). (a) Estimated mean temperature over time in Australia compared to RPANDA global paleotemperature data. (b) Estimated mean aridity index over time based on precipitation threshold (arid < 500mm/year). Abbreviations Pli = Pliocene and Ple = Pleistocene.'}

@@

```{r fig-RPANDA, fig.cap= 'Effect of aridity change on the diversification of Anilios estimated from the phylogeny. (a) The aridity index through time calculated from reconstructed paleoclimate (dotted line), and lineage through time estimated from the phylogeny (solid line). (b) Fitted diversification rates through time reconstructed from parameter estimates (λ = 1.32, α = -0.14, µ = 0.23) from the best diversification model (speciation rates vary exponentially with paleoaridity and constant extinction rate; “BEnvVar + DCST (EXPO)”). (c) Fitted diversification rates at different aridity index values calculated from parameter estimates of the previous model. Increased aridity is associated with decrease net diversification rates in this model.'}

@@

Tiatragul et al., (accepted) proposed that the earliest divergence of *Anilios* occurred around 25 Ma (28 – 22 Ma; 95% HPD), which is slightly older than 21.9 Ma (30.2 – 15.1 Ma; 95% Bayesian credibility interval) proposed by previous studies [@marinTracing2013;@vidalBlindsnake2010]. Our BioGeoBEARS analyses suggest that _Anilios_ gradually diversified in the tropical grassland (Group 1) and arid biome (Group 2) during a time when Australia was relatively wet and humid [@byrneBirth2008]. This indicates that many species found in arid biomes, particularly in Group 2, diversified in the arid biome prior to the onset of widespread aridification in the late Miocene [@byrneBirth2008].

@@

## Limitations
Although we did not detect conflicting signal between our analyses, some of our results, particularly those that estimate diversification rates (GeoHiSSE, ClaDS, and RPANDA), may be biased due to the small sample size of our _Anilios_ phylogeny [@caetanoHidden2018;@malietModel2019;@sunEstimating2020]. However, this near-complete phylogeny of _Anilios_ is the most comprehensive data set available at this time. 
We suspect there remains several undescribed species of blindsnakes in Australia and New Guinea that may belong in the genus _Anilios_. Future studies should endeavour to include more samples per species to increase the chance of detecting cryptic OTUs. We also wish to highlight that obtaining genetic data from _A. erycinus_ should be a priority as this will allow us to place the species among other _Anilios_ and provide a more comprehensive picture of the evolution of this genus.
